A highly-available move operation for replicated trees

### 2019-06-10 paper/move-op.tex
paper: start describing algorithm

In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, then the move operation must be rejected.
However, in a replicated setting a new problem arises: different replicas may perform operations that individually are safe, but whose combination leads to a cycle.

@@

Najafzadeh et al. propose two solutions to this problem: either to duplicate tree nodes, as in Figure~\ref{fig:move-cycle}b, or to execute a synchronous locking protocol that prevents two move operations concurrently modifying the same part of the tree.
The downside of a locking protocol is that the move operation is no longer highly available in the presence of network partitions, since it must wait for communication with other replicas.

While these solutions are valid, the authors go on to claim that these are the only possible options, asserting that ``no file system can support an unsynchronised move without anomalies, such as loss or duplication''~\cite{Najafzadeh:2018bw}.
We revise that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisified (in particular, it never introduces cycles), and it never duplicates any tree nodes.
To our knowledge, our algorithm is the first to provide all of these properties simultaneously.
We give a precise specification of our algorithm's consistency properties in Section~\ref{sec:proof}.

% - No Synchronisation or consensus required. Available in the sense of CAP
% - Rely only on uniqueness of node IDs, e.g. inode numbers
% - Composing JSON/XML tree by using list datatype
% - Location is Lww register. Concurrent move does not duplicate
% - Isabelle: combines the conciseness of pseudocode with the precision of mathematical notation.
% - Unlike NFS, can work offline 
% - Garbage collection: discard causally stable prefix

@@

\ interp{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymRightarrow}\isanewline

@@

\ \ {\isacartoucheopen}interp{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline

@@

\ \ {\isacartoucheopen}interp{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}logop\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline

@@

\ \ \ \ \ \ then\ redo{\isacharunderscore}op\ logop\ {\isacharparenleft}interp{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}ops{\isacharcomma}\ undo{\isacharunderscore}op\ {\isacharparenleft}logop{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}{\isacharparenright}{\isacharparenright}\isanewline

@@

\ interp{\isacharunderscore}ops\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ list\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
\ \ {\isacartoucheopen}interp{\isacharunderscore}ops\ ops\ {\isasymequiv}\ foldl\ {\isacharparenleft}{\isasymlambda}state\ oper{\isachardot}\ interp{\isacharunderscore}op\ oper\ state{\isacharparenright}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ {\isacharbraceleft}{\isacharbraceright}{\isacharparenright}\ ops{\isacartoucheclose}\isanewline

@@

@@

>>>>

In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, then the move operation must be rejected.
However, a new problem arises in a replicated setting: different replicas may perform operations that individually are safe, but whose combination leads to a cycle.

@@

Najafzadeh et al. propose two solutions to this problem: either to duplicate tree nodes, as in Figure~\ref{fig:move-cycle}b, or to execute a synchronous locking protocol that prevents two move operations from concurrently modifying the same part of the tree.
The downside of a locking protocol is that the move operation is no longer highly available in the presence of network partitions, since it must wait for communication with other replicas.

While these solutions are valid, the authors go on to claim that ``no file system can support an unsynchronised move without anomalies, such as loss or duplication''~\cite{Najafzadeh:2018bw}.
We revisit that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisified (in particular, it never introduces cycles), and it never duplicates or loses any tree nodes.
To our knowledge, our algorithm is the first to provide all of these properties simultaneously.
We give a precise specification of our algorithm's consistency properties in Section~\ref{sec:proof}.

@@

\ apply{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymRightarrow}\isanewline

@@

\ \ {\isacartoucheopen}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline

@@

\ \ {\isacartoucheopen}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}logop\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline

@@

\ \ \ \ \ \ then\ redo{\isacharunderscore}op\ logop\ {\isacharparenleft}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}ops{\isacharcomma}\ undo{\isacharunderscore}op\ {\isacharparenleft}logop{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}{\isacharparenright}{\isacharparenright}\isanewline

@@

\ apply{\isacharunderscore}ops\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ list\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
\ \ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isasymequiv}\ foldl\ {\isacharparenleft}{\isasymlambda}state\ oper{\isachardot}\ apply{\isacharunderscore}op\ oper\ state{\isacharparenright}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ {\isacharbraceleft}{\isacharbraceright}{\isacharparenright}\ ops{\isacartoucheclose}\isanewline

@@

We now introduce our algorithm for a replicated tree data structure that supports a move operation.
We model each replica as a state machine that transitions from one state to the next by applying an operation.
The algorithm is executed independently on each replica with no shared memory between replicas.

When the user makes a change to the tree, that change is expressed by generating an operation on their local replica.
We assume the existence of a messaging middleware that copies operations between replicas, so that eventually every operation is applied on every replica, provided that network partitions are of a finite duration.
This middleware may deliver operations in any order.
Operations received from a remote replica are applied in exactly the same way as operations generated by a local user.

The key consistency property of our algorithm is \emph{convergence}: that is, whenever any two replicas have applied the same set of operations, then they must be in the same state---even if the operations were applied in a different order on different replicas.
We prove this in Section~\ref{sec:proof} by showing that with our algorithm, applying operations is commutative.

Figure~\ref{fig:code} gives the full source code for our algorithm, implemented using the Isabelle/HOL language~\cite{DBLP:conf/tphol/WenzelPN08}.
We choose this language because it combines the conciseness of pseudocode with the precision of mathematical notation.
It supports formal reasoning, allowing us to prove the correctness of the algorithm (Section~\ref{sec:proof}), and it also allows executable code to be exported to common functional programming languages such as Scala, Haskell, OCaml, or Standard ML (Section~\ref{sec:evaluation}).

In this section we walk through the code in Figure~\ref{fig:code} step by step, explaining features of the Isabelle/HOL syntax as we encounter them.
For additional background on the language, standard documentation is available~\cite{DBLP:books/sp/NipkowK14}.

\subsection{Operations and trees}\label{sec:ops-trees}

We define only one kind of operation: \isa{Move t p m c} (Figure~\ref{fig:code}, lines 1--5).
A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be instantiated with arbitrary datatypes; our only requirement is that timestamps \isa{'t} are globally unique and have a total order.
For example, Lamport timestamps~\cite{Lamport:1978jq} meet these requirements.

The meaning of an operation \isa{Move t p m c} is that at time \isa{t}, the child node \isa{c} is moved to be a child of parent node \isa{p}.
The operation does not specify the old location of \isa{c}; the algorithm simply ``steals'' \isa{c} from wherever it is currently located in the tree, and moves it to \isa{p}.
If \isa{c} does not currently exist in the tree, it is created as a child of \isa{p}.
For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
Moreover, we do not define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.

The metadata field \isa{m} in a move operation allows additional information to be associated with the parent-child relationship of \isa{p} and \isa{c}.
For example, in a filesystem, the parent and child would be the inodes of a directory and a file within it, respectively, and the metadata would contain the filename of the child.
Thus, a file with inode \isa{c} can be renamed by performing a \isa{Move t p m c}, where the new parent directory \isa{p} is the unchanged inode of the existing parent directory, but the metadata \isa{m} contains the new filename.

When users want to make changes to the tree on their local replica, they generate new \isa{Move t p m c} operations for these changes, and apply these operations to their local replica.
The timestamp \isa{t} of each operation must be globally unique; the other fields \isa{p}, \isa{m} and \isa{c} can be chosen freely.

We can now represent the tree as a set of \isa{(parent, meta, child)} triples, denoted in Isabelle as \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set}.
When we have \isa{(p, m, c) {\isasymin} tree}, that means \isa{c} is a child of \isa{p} in the tree, with associated metadata \isa{m}.
Given a \isa{tree}, we can move the child \isa{c} to a new parent \isa{p}, with associated metadata \isa{m}, as follows:
\begin{quote}
\isa{\{(p', m', c') {\isasymin} tree. c' {\isasymnoteq} c\} {\isasymunion} \{(p, m, c)\}}
\end{quote}
That is, we remove any existing parent-child relationship for \isa{c} from the set \isa{tree}, and then add \isa{\{(p, m, c)\}} to represent the new parent-child relationship.
This expression appears on lines 31 and 36 of Figure~\ref{fig:code}, as we shall explain shortly.

\subsection{Replica state and operation log}

In order to correctly apply move operations, a replica needs to maintain not only the current state of the tree, but also an \emph{operation log}.
The log is a linked list of \isa{LogMove} records in descending timestamp order.
A \isa{LogMove t oldp p m c} record (lines 7--12) is similar to a record \isa{Move t p m c}; the only difference is that the \isa{LogMove} has an additional field \isa{oldp} of type \isa{('n {\isasymtimes} 'm) option}.
This \isa{option} type means the field can either take the value \isa{None} (equivalent to null), or a pair of an inode and a metadata field.

When a replica applies a \isa{Move} operation to its tree, it also records a corresponding \isa{LogMove} operation in its log.
The \isa{t}, \isa{p}, \isa{m} and \isa{c} fields are taken directly from the \isa{Move} record, while the \isa{oldp} field is filled in based on the state of the tree before the move.
If \isa{c} did not exist in the tree prior to the move, \isa{oldp} is set to \isa{None}.
Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exist \isa{p'}, \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
This logic is performed by the \isa{get\_parent} function (lines 16--20).

In the first line of \isa{get\_parent}, the expression between \isa{::} and \textbf{where} is the type signature of the function, in this case:
\begin{quote}
\isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set {\isasymRightarrow} 'n {\isasymRightarrow} ('n {\isasymtimes} 'm) option}
\end{quote}
This signature denotes a function that takes two arguments: a tree \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} and a node \isa{'n}.
It then returns a \isa{('n {\isasymtimes} 'm) option}.
\isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots''.

In line 14 we define the datatype for the state of a replica: a pair \isa{(log, tree)} where \isa{log} is a list of \isa{LogMove} records, and \isa{tree} is a set of \isa{(parent, meta, child)} triples as before.

\subsection{Applying operations}

% - No Synchronisation or consensus required. Available in the sense of CAP
% - Rely only on uniqueness of node IDs, e.g. inode numbers
% - Composing JSON/XML tree by using list datatype
% - Location is Lww register. Concurrent move does not duplicate
% - Unlike NFS, can work offline 
% - Garbage collection: discard causally stable prefix

@@

\section{Evaluation}\label{sec:evaluation}
